---
title: "STAT650__fordgobike"
output:
  pdf_document: default
  word_document: default
  html_notebook: default
auther: Mulati Patiguli
---
**Prof. Eric A. Suess**

# Question1: Explain what the GBFS is?

Answer:  GBFS ,which is the abbreviaion of General Bikeshare Feed Specification , supports publicly available station bike/dock availability information and does not require the use of an API token. Developers are encouraged to review the GBFS documentation available on Github as they determine how to incorporate BCycle system information into their applications and websites and it is the open data standard for bikeshare and also  makes real-time data publicly available online in a uniform format so that maps and transportation based apps can easily incorporate this data into their platforms.

# Question 2: Explain any difficulties you encountered getting the code to work.

Answer: 1. When I try to download the each file , i need to add the "lib" before "curl" like typed as "libcurl ". 
        2. I am also having trouble with the setting my own directory  , after this i learned how to set directory by my own every time i work with some files .
        3.I found out that some date of birth were inaccurate and found out there were some outliers regarding to age variable .and also i need to remove these outliers. 
        4.Data type of age changed from int to char and had trouble with changing them while calculating the mean.
        
    

```{r}
library(tidyverse)
library(tictoc)
library(ggmap)
library(skimr)
library(lubridate)
library(forcats)
```

Set working directory.

```{r}
setwd("C:/Users/Patime/Desktop/midterm/data")
```
 
 Download the files into the data directory.  First one is not zipped, the remaining are zipped.

```{r}
URL <- "https://s3.amazonaws.com/fordgobike-data/2017-fordgobike-tripdata.csv"
download.file(URL, destfile = "./data/2017-fordgobike-tripdata.csv", method="libcurl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201801-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201801-fordgobike-tripdata.csv.zip", method="libcurl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201802-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201802-fordgobike-tripdata.csv.zip", method="libcurl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201803-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201803-fordgobike-tripdata.csv.zip", method="libcurl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201804-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201804-fordgobike-tripdata.csv.zip", method="libcurl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201805-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201805-fordgobike-tripdata.csv.zip", method="libcurl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201806-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201806-fordgobike-tripdata.csv.zip", method="libcurl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201807-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201807-fordgobike-tripdata.csv.zip", method="libcurl")
URL <- "https://s3.amazonaws.com/fordgobike-data/201808-fordgobike-tripdata.csv.zip"
download.file(URL, destfile = "./data/201808-fordgobike-tripdata.csv.zip", method="libcurl")
```

Loop over the one value in the url and filename that changes.
```{r, eval=FALSE}

URL <- "https://s3.amazonaws.com/fordgobike-data/2017-fordgobike-tripdata.csv"
download.file(URL, destfile = "./data/2017-fordgobike-tripdata.csv", method="libcurl")

for (i in 1:7) {
URL <- paste0("https://s3.amazonaws.com/fordgobike-data/20180",i,"-fordgobike-tripdata.csv.zip")
download.file(URL, destfile = paste0("./data/20180",i,"-fordgobike-tripdata.csv.zip"), method="curl")
}
```

Unzip downloaded files.

```{r}
unzip("./data/201801-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201802-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201803-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201804-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201805-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201806-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201807-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201808-fordgobike-tripdata.csv.zip",exdir="./data")
```

Clean up data directory.

```{r}
fn <- "./data/201801-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201802-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201803-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201804-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201805-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201806-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201807-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201808-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
```

Read the.csv files

```{r message=FALSE}
fordgobike2017 <- read_csv(file="./data/2017-fordgobike-tripdata.csv")
fordgobike201801 <- read_csv(file="./data/201801-fordgobike-tripdata.csv")
fordgobike201802 <- read_csv(file="./data/201802-fordgobike-tripdata.csv")
fordgobike201803 <- read_csv(file="./data/201803-fordgobike-tripdata.csv")
fordgobike201804 <- read_csv(file="./data/201804-fordgobike-tripdata.csv")
fordgobike201805 <- read_csv(file="./data/201805-fordgobike-tripdata.csv")
fordgobike201806 <- read_csv(file="./data/201806-fordgobike-tripdata.csv")
fordgobike201807 <- read_csv(file="./data/201807-fordgobike-tripdata.csv")
fordgobike201808 <- read_csv(file="./data/201808-fordgobike-tripdata.csv")
```

Check the head() and tail() of the data.frames that are loaded.

```{r}
head(fordgobike2017) 
head(fordgobike201801)
head(fordgobike201802)
head(fordgobike201803)
head(fordgobike201804)
head(fordgobike201805)
head(fordgobike201806)
head(fordgobike201807)
head(fordgobike201808)
```

```{r}
tail(fordgobike2017)
tail(fordgobike201801)
tail(fordgobike201802)
tail(fordgobike201803)
tail(fordgobike201804)
tail(fordgobike201805)
tail(fordgobike201806)
tail(fordgobike201807)
tail(fordgobike201808)
```
by dim()function , we can get total number of row observations which is 519700 and total number column variables  which is 15 in fordgobike2017 data frame.

```{r}
dim(fordgobike2017)

fordgobike2017 %>% count()

```

 
 view total number of row observations and total number column variables  which are from Jan of 2018 to Aug of 2018 in fordgobike2018 data frame.

```{r}

fordgobike201801 %>% count()
fordgobike201802 %>% count()
fordgobike201803 %>% count()
fordgobike201804 %>% count()
fordgobike201805 %>% count()
fordgobike201806 %>% count()
fordgobike201807 %>% count()

```
by glimpse()function , we can see that start_station_id data types for below months are chr .
        
```{r}
glimpse(fordgobike201805)
glimpse(fordgobike201806)
glimpse(fordgobike201807)




```


# Question 3: The analysis is to work with the 2017 and 2018 data. How many bike rentals were there in 2017? How many bike rentals were there in 2018? How many bike rentals have there been since the beginning of Ford GoBikes?

Answer: 
Number of Bike rentals in  2017 was 272
Number of Bike rentals in  2018 was 962
Number of Bike rentals since the beginning of Ford GoBikes = 272+962=1234



# Question 4: There is a part of the code that uses the as.integer() function for some reason. Explain what this function is being used for in the code.

Answer: Data type we used for Start_Station_id, end_station_id variables in the data frame, which are fordgobike201806, fordgobike201807 and fordgobike201806 data frame ,are character type.So we need to convert this kind of type to integer type by using as.integer()function. we use this function in order to match data type with other data frame. 
and by the way , if we use bind()function ,combine ()function to combine the data frames with different data type variables ,we will encounter an error . So before we use any combining function , we have to be sure our variable with the same data type .



Converting the start_station_id data type to integer for forgobike201806, forgobike201807, and forgobike201808 to match with other data. Combine all the multiple 2018 data set into single data set
```{r}
fordgobike201806 <- fordgobike201806 %>%
        mutate(start_station_id = as.integer(start_station_id),
               end_station_id= as.integer(end_station_id) )   

fordgobike201807 <- fordgobike201807 %>%
        mutate(start_station_id = as.integer(start_station_id),
               end_station_id= as.integer(end_station_id) )

fordgobike2018 <- bind_rows(fordgobike201801, fordgobike201802, fordgobike201803, fordgobike201804,
        fordgobike201805, fordgobike201806, fordgobike201807)

glimpse(fordgobike2018)
```

we combined data frame fordgobike2017 and fordgobike2018 .


```{r}
dim(fordgobike2017)

fordgobike2017 %>% count()

nrow(fordgobike201801) + nrow(fordgobike201802) + nrow(fordgobike201803) + nrow(fordgobike201804)

dim(fordgobike2018)

fordgobike2018 %>% count()

fordgobike <- bind_rows(fordgobike2017, fordgobike2018)

dim(fordgobike)




```
# Question 5: How is the Age variable created? Are there any outliers in the data? How many outliers have you removed? State what you think is a good cut off is to remove any outliers.

Answer:
if we look at  the fordgobike data frame,we can figure out that member_ Birth_year variable's value is given. So before create the new Age variable to calculate the age, we need to use mutate()function to create a new variable which the age and also this variable is created by subtracting the current year from the given variable (member_birth_date) .

if we use geom_histogram() function with filter()function on fordgobike data ,we can see the outliers. from histogram below, we can see  that the data is showing a highly right skewed pattern. however we infer that there exests an outlier.
Major portion of data lies between age range between 18 and 80 is majority proportion of data .except of this , there still showing some age group which are greater than above age 80.so we need to remove these outliers or replace them with NA value.

So I removed anything beyond 80 yrs which is considered as outliers and I replaced with NA values. 


add another new variable Age by using mutate ()function
```{r}
fordgobike %>% count()

dim(fordgobike)

fordgobike <- fordgobike %>% mutate(age = 2018 - member_birth_year)

fordgobike %>% count() 

dim(fordgobike)

fordgobike <- fordgobike %>% mutate(year=year(start_time), month=month(start_time), day=day(start_time) )


```



```{r warning=FALSE}
fordgobike %>% group_by( age ) %>% count()

fordgobike %>% group_by( age ) %>% summary()

skim(fordgobike)

fordgobike %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age <= 100) %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age > 100) %>% ggplot(aes(x=age)) + geom_histogram()
```


```{r}
fordgobike %>% group_by( member_gender, age ) %>% count()

fordgobike %>% ggplot(aes(x=age, class=member_gender)) + geom_histogram()

fordgobike %>% ggplot(aes(x=age, class=member_gender)) + geom_histogram(aes(y=..density..))

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age)) + geom_histogram()
```
Replacing the Age > 80 yrs with NA value. 

```{r}
Fordgobike_outlier_NA<-fordgobike %>% mutate(age=replace(age,age>80,"NA"))
Fordgobike_outlier_NA_Only<-Fordgobike_outlier_NA %>% filter(age=="NA") %>% count()
glimpse(Fordgobike_outlier_NA)
Fordgobike_outlier_NA %>% ggplot(aes(x=age)) + geom_histogram(stat = "count",binwidth = 30)
```
# Question 6: In 2018, what month had the highest number of riders? What month had the lowest number of riders? Interpret any seasonal patterns.

Answer: from below results, July  had the highest number of riders which is 199222 .and the January month had the lowest number of riders which is 94802.

if we interpret these two results with seasonal pattern , we can say because on July month, most of the people are willing to go outside and they will be more active ,so total rides would peak to the highest number and its number went down as the weather changes .In summer season like July ,June was first and second highest number .but in  Winters ,there was showing less bike rides ,since winter is cold ,it is reasonabale to see less bike riders outside .therefore total bike rides follows to the seasonal pattern .it increases from winter to spring and reaches peak in July . 



Filtering the 2018 data from the combined and outlier removed/NA data
```{r}
fordgobike_outlier_NA_2018 <- Fordgobike_outlier_NA %>% filter(year==2018)
Riders<-fordgobike_outlier_NA_2018  %>% group_by(month) %>% summarise(total_rides=n()) %>% arrange(desc(total_rides))
Riders
```


```{r}
fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age, color=member_gender)) + 
  geom_histogram(position="identity") + 
  facet_grid(member_gender ~ .)

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age, color=member_gender)) + 
  geom_histogram(aes(y=..density..),position="identity") + 
  facet_grid(member_gender ~ .)

```

# Question 7: What start station had the highest number of rides? That is, which start station was used most to start rides?

Answer: Start_station_ID 15 had highest number of rides. That is San Francisco Ferry Building  was used most to start rides.




Group by station id and name 
```{r}
station<-Fordgobike_outlier_NA %>% group_by(start_station_id,start_station_name) %>% summarise(A=n())  %>% arrange(desc(A))
station
                                                                                      
```
# Question 8: What was the Age of the youngest rider? What was the Age of the oldest rider, after removing the outliers? What was the mean Age of the rider? What was the mean Age of the Female riders? What was the mean Age of the Male riders?

Answer: 
  Age of the youngest rider ------18 yrs
  Age of the oldest rider ------ 80 yrs
  mean age of rider ------ 36.54
  Mean age of Male rider - -----36.54
  Mean age of Female rider - 34.69
```{r}
Fordgobike_outlier_NA %>% group_by(age) %>% count() %>% arrange(desc(age))

Fordgobike_outlier_NA$age<-as.integer(Fordgobike_outlier_NA$age)

Fordgobike_outlier_NA %>% na.omit(age)

glimpse(Fordgobike_outlier_NA)

Fordgobike_outlier_NA %>% group_by(age) %>% summary()


```
```{r}
male<-Fordgobike_outlier_NA %>% filter(member_gender=="Male") %>% group_by(age) %>% summary()
male
Female<-Fordgobike_outlier_NA %>% filter(member_gender=="Female") %>% group_by(age) %>% summary()
Female
```


# Question 9: Using the Amelia package and the missmap() function determine the rate of missing data in the month of June.

Answer: 1% of data is missing and 99% of data observed in the month of june.


To calculate missing observation in the month of June
```{r}


fordgobike_june<-Fordgobike_outlier_NA %>% filter(month==6)
fordgobike_june
missmap(fordgobike_june)

```

#Question 10: What Type of rider uses the Ford goBikes more? Subscribers or Customers?


Based on the below results, we can find  that subscriber group are more than customers group.


To find out no of subscribers and customer user types

```{r}
Fordgobike_outlier_NA %>% filter(user_type=="Subscriber") %>% count()

```
```{r}
Fordgobike_outlier_NA %>% filter(user_type=="Customer") %>% count()
```